An agent-based model for modal shift in public transport

### 2021-07-23 Response_to_reviewers.tex

\documentclass[10pt]{article}

\usepackage[margin=2cm]{geometry}

\begin{document}

\section*{Response to reviewers}

\subsection*{Reviewer 1}

% The paper introduces an agent-based model aimed at understanding the impact of behavioural parameters on modal shift. The model is applied on a case study based on a stated preference survey for a segment of Paris suburban train network. The research contributions are clearly stated. The methodology is appropriate. The main concern is with sample size; 48 users seem rather short. The paper misses a separate chapter dedicated to conclusions. The formatting rules of the template should be verified and implemented.

\noindent\textit{The main concern is with sample size; 48 users seem rather short.}

\noindent$\rightarrow$ The empirical study was a small pilot that could not be continued - thus the modeling approach to still be able to exploit it. This point was emphasised again in the final paper.

\medskip

\noindent\textit{The paper misses a separate chapter dedicated to conclusions.}

\noindent$\rightarrow$ A conclusion section was added.

\medskip

\noindent\textit{The formatting rules of the template should be verified and implemented.}

\noindent$\rightarrow$ The EWGT 2021 Latex template provided was used to format the final paper.

\bigskip

\subsection*{Reviewer 2}

% The authors propose an agent-based model for modal shift in public transport. The subject of the paper is interesting with evident potential practical implications.
%The weakest part of the article is that the model parameters are built on a very small sample of passengers. However, this limitation is clearly assumed by the authors.
%Despite realizing that the advantage of the approach is the simplicity, the methodology section is not clear in some parts. For example:
% - On which the authors rely to ensure “that perceived congestion given by the current number of user waiting” must be normalized by the platform capacity?
% - How is cross demand elasticity of other modes considered? (e.g., in terms of ticketing, frequency and information quality for other modes?)
%- Is the existence of a mobility service that covers all the alternatives assumed?
%- How the perceived value of travel time varies regarding the waiting time at the stop and the journey time itself? Is the cost of the loss of comfort related to crowding (e.g., pandemic effects) possible to be simulated?

%Considering, the potential applicability of the model based on open source and the honesty about the level of development (still premature), I recommend its presentation at the EWGT. Before being accepted, the article should be improved by clarifying some methodological aspects, adding conclusions, and improving formatting according EWGT template rules.

\noindent\textit{On which the authors rely to ensure ``that perceived congestion given by the current number of user waiting'' must be normalized by the platform capacity?}

\noindent$\rightarrow$ This normalisation is in our case a convenience rescaling to have comparable scales for each discrete choice parameter, and does not impact results as we have a single boarding station with fixed capacity in our model. Measuring effective perceived congestion in multiple stations and users is a difficult problem - this was emphasised in the paper discussion.

\medskip

\noindent\textit{How is cross demand elasticity of other modes considered? (e.g., in terms of ticketing, frequency and information quality for other modes?)}

\noindent$\rightarrow$ These aspects are implicitly included in the fixed coefficients for modal shift but are not explicitly included in the model. A model refinement could consider such developments - this was added in the model description and paper discussion.

\medskip

\noindent\textit{Is the existence of a mobility service that covers all the alternatives assumed?}

\noindent$\rightarrow$ We do not explicitly assume a unified mobility service (RER/subway are a distinct service from bikes and taxis for example) nor a single application synthesising all the information available on all alternatives. The exploration of this aspect was added as a potential development.

\medskip

\noindent\textit{How the perceived value of travel time varies regarding the waiting time at the stop and the journey time itself? Is the cost of the loss of comfort related to crowding (e.g., pandemic effects) possible to be simulated?}

\noindent$\rightarrow$ In our case, perceived value of travel time is simply the total travel time plus the expected waiting time given by current information on the platform. Exploring diverse functions and aspects for this disutility such as crowding is indeed an interesting development, which was added in the discussion.

\end{document}

### 2021-07-23 paper.tex
Update on Overleaf.

%Notre but est de déterminer si l’accès à l’information sur le temps d’attente lors d’une perturbation pour les usagers est utile ou non. Pour commencer nous établissons un périmètre d’étude. Dans notre cas nous nous intéressons au RER A, il nous faut cependant réduire notre périmètre pour nous concentrer sur une zone plus précise. Nous souhaitons vérifier l’impact sur le réseau, pour cela nous choisissons donc des stations qui soient suffisamment maillées entre elles pour pouvoir dessiner un “mini-réseau”. Le but est de construire notre étude pour un trajet s’effectuant d’un point A à un point B. De plus, afin de s’assurer que les usagers effectuent le trajet qui nous intéresse, et par simplification de notre “mini-réseau”, nous devons choisir deux stations proches l’une de l’autre. Nous étudierons donc ici le report modal (ou le reroutage) en cas de situation perturbée sur la ligne du RER.

@@

We surveyed a total of $N=48$ users, among which a subsample of $N=27$ were regular users for which the full questionnaire was given. Main results which can inform our model context are shown in Fig.~\ref{fig:fig1}. We first confirm that disruptions are indeed frequent on the line, as around 50\% of users experience a disruption at least twice a month. We then find that under no information, user behavior is heterogeneous and a large part shifts to alternative itineraries. Finally, when users are informed of a 10 minutes disruption, on the contrary more than 75\% decide to wait on place. Therefore, the model must include various modal shifts, but also take into account how perceived wait time will influence user choice.

This survey was designed in an exploratory manner and could not be continued on a larger scale for various practical reasons. Our sample size thus remains relatively small and this part of the study must be considered as qualitative. Significantly larger sample sizes are needed to obtain robust inferential results, for example to forecast travel demand \citep{horvath2015estimation} or to estimate discrete choice parameters \citep{bliemer2009efficiency}. In our case, we consider the simulation model as a tool to go beyond such data limitations and still study stylised behavioural processes.

@@

\begin{enumerate}
	\item New users enters the network: for each mode, a fixed arrival rate $\lambda_i$ is used for a random Poisson law draw giving the number of new users entering the segment on this mode. Users entering the main RER mode are set on the platform as waiting, while users on other modes are queued at the start of the segment.
	\item Users waiting for a RER train on the platform evaluate a discrete choice utility difference (between waiting and shifting to an other mode) given by 
	\begin{equation} 
	\Delta U = \beta_c \cdot c + \beta_{\tau} \tau
	\end{equation}
	where $c$ is perceived congestion given by the current number of user waiting (normalised by platform capacity) and $\tau$ is perceived time proportional to user current travel time (time since departure) and current waiting time between trains. $\beta_c$ and $\beta_{\tau}$ are thus behavioural parameters capturing the influence of perceived time and congestion on user behaviour.
	\item These users can then switch mode with the corresponding discrete choice probability probability given by 
	\begin{equation}
	p = 1 / (1 + \exp \Delta U)
	\end{equation}
	and with fixed nested probabilities for the choice of the other mode once a shift has been decided;
	\item Next train possibly enter station (once the time interval $I$ between trains has elapsed) and users board at a given speed and given train capacity. Train capacity $C$ captures congestion upstream this segment, and is given as a maximal number of users which can board a train. In the meantime, users alight at the end of the segment if a train has arrived at the terminus station.
	\item Train queue on the segment is simulated, i.e. each train is advanced at its maximal speed, if the next slot is not occupied by a preceding train.
	\item Other modes are simulated as queues with a given user capacity: users enter the queue if it contains less users than its capacity; users in the queue are advanced according to the mode speed; arrived users are removed from the queue.
\end{enumerate}

@@

@@

%%%%%%%%%%%%%
\begin{figure}%[t]\vspace*{4pt}
\begin{mdframed}
\centerline{\includegraphics[width=\linewidth]{figures/Fig2.png}}
\end{mdframed}
\caption{Model interface. Main varying parameters are included as sliders, while mode queues are visualised in a spatialised manner and as a load plot in time. Various numerical indicators are provided.\label{fig:fig2}}
\end{figure}
%%%%%%%%%%%%

@@

Several developments can furthermore be considered. The behavioural model is rather simple, and as the implementation is generic, other functions or more complicated processes can be tested. The variables we used in the discrete choice utility are only proxies of perceived information, and how to account for the information effectively received by the users depending on the context remains an open question. Besides, the possible mitigation of congestion through interventions acting on the provided information may raise ethical issues - how to ensure an overall optimisation while remaining fair to all users. Finally, a more data-driven version of the model, including estimated discrete choice parameters, would be an interesting path to explore, assuming that a further survey and data collection is possible.

@@

%\vspace{-0.5cm}

>>>>

@@

We surveyed a total of $N=48$ users, among which a sub-sample of $N=27$ were regular users for which the full questionnaire was given. Surveys were realised on peak hours (between 7am and 9am) of weekdays in January 2021, on the platform of the \emph{Etoile} station. The exact survey which was used (in French) is available on the open repository of the project (see model implementation below).

Main results which can inform our model context are shown in Fig.~\ref{fig:fig1}. We first confirm that disruptions are indeed frequent on the line, as around 50\% of users experience a disruption at least twice a month. We then find that under no information, user behavior is heterogeneous and a large part shifts to alternative itineraries. Finally, when users are informed of a 10 minutes disruption, on the contrary more than 75\% decide to wait on place. Therefore, the model must include various modal shifts, but also take into account how perceived wait time will influence user choice.

This survey was designed in an exploratory manner as a pilot study, and could not be continued on a larger scale for various practical reasons. Our sample size thus remains relatively small and this part of the study must be considered as qualitative. Significantly larger sample sizes are needed to obtain robust inferential results, for example to forecast travel demand \citep{horvath2015estimation} or to estimate discrete choice parameters \citep{bliemer2009efficiency}. In our case, we consider the simulation model as a tool to overcome such data limitations and still study stylised behavioural processes using qualitative insights from our survey data.

@@

\begin{enumerate}
	\item New users enters the network: for each mode, a fixed arrival rate $\lambda_i$ is used for a random Poisson law draw giving the number of new users entering the segment on this mode. Users entering the main RER mode are set on the platform as waiting, while users on other modes are queued at the start of the segment.
	\item Users waiting for a RER train on the platform evaluate a discrete choice utility difference (between waiting and shifting to an other mode) given by 
	\begin{equation} 
	\Delta U = \beta_c \cdot c + \beta_{\tau} \tau
	\end{equation}
	where $c$ is perceived congestion given by the current number of user waiting normalised by platform capacity (which corresponds in our case only to a parameter rescaling as we consider a single station); $\tau$ is perceived time that we assume proportional to user current travel time (time since departure of the user trip) and current waiting time between trains. $\beta_c$ and $\beta_{\tau}$ are thus behavioural parameters capturing the influence of perceived time and congestion on user behaviour. Note that a model generalisation with a broader scope (more stations for example) will have to introduce more accurate proxies for this processes or other behavioural aspects.
	\item These users can then switch mode with the corresponding discrete choice probability probability given by 
	\begin{equation}
	p = 1 / (1 + \exp \Delta U)
	\end{equation}
	and with fixed nested probabilities for the choice of the other mode once a shift has been decided;
	\item Next train possibly enter station (once the time interval $I$ between trains has elapsed) and users board at a given speed and given train capacity. Train capacity $C$ captures congestion upstream this segment, and is given as a maximal number of users which can board a train. In the meantime, users alight at the end of the segment if a train has arrived at the terminus station.
	\item Train queue on the segment is simulated, i.e. each train is advanced at its maximal speed, if the next slot is not occupied by a preceding train.
	\item Other modes are simulated as queues with a given user capacity: users enter the queue if it contains less users than its capacity; users in the queue are advanced according to the mode speed; arrived users are removed from the queue.
\end{enumerate}

@@

The discrete choice function does not include control variables such as differences between modes in ticket price, level of service, quality of information. These are implicitly taken into account in our model through the empirical values of modal shift observed in the survey, and could be included in a refined version of the model.

@@

%%%%%%%%%%%%%
\begin{figure}%[t]\vspace*{4pt}
\begin{mdframed}
\centerline{\includegraphics[width=\linewidth]{figures/Fig2.png}}
\end{mdframed}
\caption{Model interface. Main varying parameters are included as sliders, while mode queues are visualised in a spatialised manner and as a plot showing the load for each mode in time. Various numerical indicators are provided.\label{fig:fig2}}
\end{figure}
%%%%%%%%%%%%

@@

Several developments can furthermore be considered. The behavioural model is rather simple, and as the implementation is generic, other functions or more complicated processes can be tested. The variables we used in the discrete choice utility are only proxies of perceived information, and how to account for the information effectively received by the users depending on the context remains an open question \citep{gao2018trip}. The disutility due to perceived travel time is in our case simply the sum of previous travel time and announced waiting time: studying other functions and aspects is an other interesting development of the model. For example, including congestion within the trains may be crucial regarding recent health concerns linked to crowding in closed spaces \citep{raimbault2021estimating}.

Besides, the possible mitigation of congestion through interventions acting on the provided information may raise ethical issues - how to ensure an overall optimisation while remaining fair to all users: it was shown in other modes such as car travel that inequalities in congestion may rise from path shift due to live information \citep{cabannes2017impact}. We also assumed on this issue of access to information that users did not necessarily had access to some synthetic information on all alternatives (with a web application for example) and that there was no unified mobility service. Exploring the impact of such aspects on behavioural and congestion patterns is also a potential refinement of the model.

Finally, a more data-driven version of the model, including estimated discrete choice parameters, would be an interesting path to explore, assuming that a further survey and data collection is possible. This may also imply to include control variables in discrete choices utilities to account for differences between modes for example in terms of price or level of service.

\section{Conclusion}

The study of modal shift following a disruption in public transport networks is an important aspect to ensure resilience and a high level of service. We introduced in this paper the basis for a simple agent-based model focused on user decision based on perceived congestion and waiting time. We showed how the model can be used to explore the interaction between different behavioural components of users, and to optimise the congestion in different modes. Our model is open source and aimed at being extended and applied to other case studies. Systematic model exploration and validation should provide both practical results in terms of transport management, but also more general results on generic and specific processes in modal shift.

@@
